---
title: "General Social Survey"
author: "Erika Mondragon"
date: "Fall 2023"
output: 
  html_document:
    number_sections: true
    toc: true
---


```{r include=FALSE,echo=FALSE}
require(tigerstats)
require(tidyverse)
require(foreign)
source("GrabData.R")
```




#  Question 1

<!--In this section you explain what you are trying to show.  Where did the data come from?  What is the research or other question you are trying to answer?!-->

The data in the dataframe GSSdata is from the 2018 General Social Survey. The first blocks of R-code has selected down a subset of the data to just 16 variables. It has further removed unwanted factor levels in much of the data. Examine the code in the GrabData.R file to see what it is doing. Some of the variables are categorical and others are numerical. Be sure to do a variable analysis before tackling each question.  
First question - Is opinion on the death penalty (CAPPUN) independent of gun ownership (OWNGUN)?


$H_0$: There is no relationship between gun ownership and opinion on the death penalty

$H_A$: There is a relationship between gun ownership and opinion on the death penalty

## Methods

<!--Decide on your methods:  use "variable analysis" or other appropriate descriptors.  Make sure to choose at least one graphical method and at least one numerical method.!-->

Both variables are categorical, therefore we use the Cat~Cat file

##Results

<!--Divide this section into two sub-sections:  One for your descriptive  results and one for your inferential results.!-->

### Descriptive Results

#### Graphical Descriptive Results

<!--Graphical results here.  Make sure to show your code.  Provide appropriate labels for axes, giving units if possible, and provide a good title for the graph, too.  Use the graphical results to describe the patterns if any that exist in the data as focused toward the research question!-->

```{r}
barchartGC(~OWNGUN + CAPPUN,data=GSSdata)
barchartGC(~OWNGUN + CAPPUN,data=GSSdata, type="percent")

```

This graph shows that those who own a gun are most likely to favor the death penalty , and those who do not own a gun are less likely to favor the death penalty

#### Numerical Descriptive Results

<!--Numerical results go here. Use the numerical results to describe the patterns if any that exist in the data as focused toward the research question!-->

```{r}
table1 <- xtabs(~OWNGUN + CAPPUN,data=GSSdata)
rowPerc(table1)
colPerc(table1)
```

This summary shows that those who own a gun are more likely to be on favor of the death penalty and those who don't own a gun are less likely to favor the death penalty 

### Inferential Results

<!--State hypothesis clearly.  Make sure your discussion of the inferential test covers all the aspects that the test output produces, such as test statistic, p-value etc.  Make a decision about the null hypothesis, explain the assumptions on which the selected test/procedure was based, and why the chosen procedure satisfys the assumptions and is appropriate to answer the research question!-->

```{r}
chisq.test(table1)
chisqtestGC(table1)
fisher.test(table1)
```

The chi-squared test p-value is 0.02022, is less than 0.05, therefore we reject the null hypothesis. 

The fisher exact test p-value is 0.01651, is less than 0.05. So we conclude that there is a relationship between gun relationship and opinion in capitol punishment. People that own a gun are on average 2.3 times more likely to favor capitol punishment than those who do not own a gun. 

#  Question 2

<!--In this section you explain what you are trying to show.  Where did the data come from?  What is the research or other question you are trying to answer?!-->

Is there a difference in years of education (EDUC) between the those who did or did not vote in 2012 (VOTE12)?

$H_0$: There is no difference between years of education and those who did or did not vote in 2012

$H_A$: There is no difference between years of education and those who did or did not vote in 2012

## Methods

<!--Decide on your methods:  use "variable analysis" or other appropriate descriptors.  Make sure to choose at least one graphical method and at least one numerical method.!-->

The response variable is EDUC, which is numerical variable. The predictor variable is VOTE12, which is categorical variable. The helper file that we would use is the numerical~cat file. 


##Results

<!--Divide this section into two sub-sections:  One for your descriptive  results and one for your inferential results.!-->

### Descriptive Results

#### Graphical Descriptive Results

<!--Graphical results here.  Make sure to show your code.  Provide appropriate labels for axes, giving units if possible, and provide a good title for the graph, too.  Use the graphical results to describe the patterns if any that exist in the data as focused toward the research question!-->

```{r}
plot(EDUC~VOTE12,data=GSSdata)
```

This plot shows that those who voted in 2012 have a higher level of years of education, and those people that didn't vote in 2012 have a lower years of education. 

#### Numerical Descriptive Results

<!--Numerical results go here. Use the numerical results to describe the patterns if any that exist in the data as focused toward the research question!-->

```{r}
favstats(EDUC~VOTE12,data=GSSdata)
```

This table shows that those who voted in 2012 have a higher level of years of education, and those people that didn't vote in 2012 have a lower years of education. 

### Inferential Results

<!--State hypothesis clearly.  Make sure your discussion of the inferential test covers all the aspects that the test output produces, such as test statistic, p-value etc.  Make a decision about the null hypothesis, explain the assumptions on which the selected test/procedure was based, and why the chosen procedure satisfys the assumptions and is appropriate to answer the research question!-->

```{r}
t.test(EDUC~VOTE12,data=GSSdata)
```

This t-test p-value is 0.0000001248, which is lower than 0.05, therefore we reject the null hypothesis, which is that there is no difference between years of education and those who did or did not vote in 2012. We conclude that there is a difference between years of education and those who did or did not vote in 2012

#  Question 3

<!--In this section you explain what you are trying to show.  Where did the data come from?  What is the research or other question you are trying to answer?!-->

Determine if a persons performance on the wordsum test (WORDSUM) is independent of their level of schooling (DEGREE).

$H_0$ There is no difference between the average score on the WORDSUM test and level of schooling DEGREE  
$H_A$ There is a significant difference in average score on the WORDSUM test between at least two levels of DEGREE  

## Methods

<!--Decide on your methods:  use "variable analysis" or other appropriate descriptors.  Make sure to choose at least one graphical method and at least one numerical method.!-->

The response variable is WORDSUM, which is numerical variable. The predictor variable is DEGREE, which is categorical variable. The helper file that we would use is the numerical~cat file. 

##Results

<!--Divide this section into two sub-sections:  One for your descriptive  results and one for your inferential results.!-->

### Descriptive Results

#### Graphical Descriptive Results

<!--Graphical results here.  Make sure to show your code.  Provide appropriate labels for axes, giving units if possible, and provide a good title for the graph, too.  Use the graphical results to describe the patterns if any that exist in the data as focused toward the research question!-->
```{r}
# Create a box plot
ggplot(data=GSSdata, aes(x = DEGREE, y = WORDSUM)) +
  geom_boxplot() +
  xlab("Category") +
  ylab("WORDSUM") +
  ggtitle("Boxplot of Response Variable by Category")
```

This graph shows the WORSUM scores of those with a Bachelor's degree, graduate, high school, and junior college dregree, and less than highschool level of education. 

#### Numerical Descriptive Results

<!--Numerical results go here. Use the numerical results to describe the patterns if any that exist in the data as focused toward the research question!-->
```{r}
favstats(WORDSUM~DEGREE,data=GSSdata)
```

This data shows the mean and median scores of WORDSUM test based on the level of education of a person. This data shows that those who received bachelor, graduate and junior college level of education scored higher on the WORDSUM test, who scored around 9-10,  than those in high school an LT high school who scored around 7-8

### Inferential Results

<!--State hypothesis clearly.  Make sure your discussion of the inferential test covers all the aspects that the test output produces, such as test statistic, p-value etc.  Make a decision about the null hypothesis, explain the assumptions on which the selected test/procedure was based, and why the chosen procedure satisfys the assumptions and is appropriate to answer the research question!-->

```{r}
model1 <- aov(WORDSUM~DEGREE,data=GSSdata)
summary.aov(model1)
```

The p-value is 0.00000025, which is lower than 0.05. Therefore, we reject the null hypothesis. This means that there is a significant difference in average score on the WORDSUM test between at least two levels of DEGREE. 


# Discussion and Conclusion

<!--Discussion and conclusion here.  If you found a relationship be sure to consider whether the relationship occurs because one of the variavbles causes the other, or whether they perhasps are related for some other reason.  Watch the chapter 6 videos from the GeorgeTown videos collection.!-->

In all three cases the p-value is less than 0.05, therefore we rejected each of null hypotheses of the 3 questions. 
